---
title: "Predicting NBA Rookie Success from College Performance"
author: "Muhammad Shah, Alex Li"
date: "2024-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(ggplot2)
```

## Introduction

Predictive modeling of NBA rookie success based on college performance statistics provides crucial insights for sports analytics. This analysis utilizes historical data to establish relationships between collegiate achievements and professional performance in the NBA, thereby supporting decision-making in sports management.

## Research Question

How can NBA rookie success, measured by points per game, be predicted from their college basketball statistics? This study employs statistical modeling techniques to forecast NBA outcomes from detailed collegiate performance metrics.

## Data Description

This study employs two primary datasets:

1. **College Basketball Dataset**: Consists of detailed player statistics from NCAA college basketball spanning 2009 to 2021. This dataset includes metrics such as player efficiency ratings and scoring averages. 

[Source: https://www.kaggle.com/datasets/adityak2003/college-basketball-players-20092021/data]
2. **NBA Player Statistics**: Contains performance data for NBA rookies' first season from 2009 to 2021, focusing on points per game and efficiency metrics. 

[Source: https://www.kaggle.com/datasets/raunakpandey030/nba-player-stats]

### Key Variables:
- `College_points`: Average points scored per game in the player's final college season.
- `eFG%`: Effective field goal percentage in college.
- `TS%`: True shooting percentage in college.
- `NBA_points`: Average points per game during the player's rookie NBA season.

### Data Cleaning and Merging Process
The data was meticulously cleaned to ensure consistency and relevance to the study's goals. This involved:
- Renaming columns for clarity.
- Filtering records to retain only the last year of college and the first year in the NBA to focus on the transition period.
- Removing duplicate entries and normalizing data formats for accurate merging.


```{r load-data, echo=TRUE, message=FALSE, warning=FALSE}
# Load libs
library(dplyr)
library(readr)
library(stringr)

# Reading and processing the college data
college <- read_csv("CollegeBasketballPlayers2009-2021.csv") %>%
  mutate(year = as.character(year)) %>%
  rename(College_points = pts) %>%
  arrange(desc(year)) %>%
  distinct(player_name, .keep_all = TRUE) %>%
  select(player_name, eFG, conf, TS_per, College_points, year)

# Reading and processing the NBA data
nba <- read_csv("NBA_Player_Stats.csv") %>%
  mutate(Year = as.character(Year),
         Year = str_split(Year, '-', simplify = TRUE)[, 1]) %>%
  arrange(Year) %>%
  rename(NBA_points = PTS,
         player_name = Player) %>%
  distinct(player_name, .keep_all = TRUE) %>%
  select(player_name, NBA_points)

# merging
merged_points <- inner_join(college, nba, by = "player_name")
```
```{r}
library(ggplot2)

# Histogram of NBA Points
ggplot(merged_points, aes(x = NBA_points)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(title = "Distribution of NBA Rookie Points", x = "NBA Points per Game", y = "Frequency")

# Scatter plot of College Points vs. NBA Points
ggplot(merged_points, aes(x = College_points, y = NBA_points)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Relationship between College Points and NBA Rookie Points", x = "College Points per Game", y = "NBA Points per Game")
```


